---
title: "difference_in_differences"
format: html
editor: visual
---

Imagine that a California state law was passed in 2016 that raised the minimum wage beginning in the year 2017. We are interested in what this law’s impact has been on student wages since student jobs often pay low wages.

We have the data on the average student wages at all California public universities from the year 2007 to 2017, displayed in the plot in the learning environment. We note that wages rise and fall from year to year, but we see a particularly large rise from 2016 to 2017.

It’s easy to assume the large increase was entirely due to the new minimum wage law. But what would average student wages in 2017 have looked like had the law NOT been passed?

Would wages still have increased a lot, meaning the law had little impact?
Would wages have decreased, meaning the law had an even larger impact than just the difference between 2016 and 2017 wages?
The fundamental principle of causal inference tells us that we can’t observe both situations, so we need to use other observed data as a substitute for what student wages in California would have looked like in 2017 had the law not passed.

```{r}

# import libraries
library(dplyr)
library(ggplot2)
# plot wages versus years
ca_wages <- wages %>%
  #only California schools
  filter(state == "California") %>% 
  #wages over time
  ggplot(aes(x = year, y = avg_wage)) + 
  #line plot
  geom_line()

# dashed line vertical
ca_wages +
  geom_vline(xintercept = 2016, linetype = "dashed") +
  scale_x_continuous(breaks = c(2007:2017))

```

Let’s say there was a new entertainment tax in Sydney starting in 2018. You want to find out if the tax affected movie theater ticket sales. You have data about average annual movie theater ticket sales in Sydney from 2012 through 2019 with the following variables:

```{r}

tickets <- read.csv("ticket_sales.csv")

syd_sales <- tickets %>%
  filter(city == "Sydney") %>%
  ggplot(aes(x = year, y = sales)) +
  geom_line ()

syd_sales + geom_vline(xintercept = 2018, linetype = "dashed") +
scale_x_continuous(breaks = c(2012:2019))

```

We see a decrease in ticket sales in 2018 when the tax was implemented, but it was already declining in 2017, so what portion was due to the tax?

It is impossible to know what truly would have happened in California schools if the law had never happened. However, we can look for another state whose average student wages that follow a similar pattern over time — but did not receive the same treatment. We can use the other state’s data to approximate the counterfactual. With DID, we use the trend of a control group to estimate the counterfactual trend for the treated. In other words, we use DID to estimate the average treatment effect on the treated (ATT).

In this example, a control trend might be the average student wage of universities in Washington, given that Washington did not implement a new minimum wage law. We can view these two groups in a single plot using the following code:

# Exercise

The tickets dataset also contains data on average ticket sales in Toronto. Create a line plot that shows the ticket sales for both Toronto and Sydney over the same years. Be sure to change the color and type of line by city. Does Toronto have a similar trend in ticket sales from 2012 to 2018?

```{r}

ggplot(data = tickets, aes(x = year, y = sales, color = city, linetype = city)) +
  geom_line() +
  #vertical line at 2018
  geom_vline(xintercept = 2018, linetype = "dashed") +
  #x-axis scale 2012 to 2018
  scale_x_continuous(breaks = c(2012:2018))

```


In this we use Toronto as a control for Sydney, and ticket sales rose after 2018 while falling in Sydney. We don't know however, how much that tax affected Sydney

# Use Linear Regression to calculate DID

While we were able to estimate the ATT through mean differences alone, we can also use linear regression for DID. A simple DID regression model predicts the outcome from the variables for treatment group and time, along with the interaction of treatment with time.

To simplify our output, we will first transform the state variable to a treatment indicator called treat (1 for California and 0 for Washington). Then we transform year to a time indicator called time (1 for 2017 and 0 for 2016).

```{r}

# transform state to treat
wages2$treat <- ifelse(wages2$state=="California",1,0)
# transform year to time
wages2$time <- ifelse(wages2$year==2017,1,0)


```

```{r}

#        state treat year time  avg_wage
# 1 California     1 2016    0 13.311279
# 2 California     1 2017    1 16.000000
# 3 Washington     0 2016    0  9.728146
# 4 Washington     0 2017    1 10.000000

```

To create a DID regression for our student wage data, we run a model that predicts average student wages from the treatment, time, and the interaction of treatment and time. Note that in R, treat*time is equivalent to treat + time + treat:time.

```{r}

did_mod <- lm(
  #include interaction
  avg_wage ~ treat*year,
  #use subsetted data 
  data = wages2 
)

```

When we print did_mod we get the following output. Note that the coefficient on the interaction term treat:year is exactly what we computed for the ATT by taking the difference of means. We estimate the impact of the minimum wage law on California student wages to be an increase of about $2.42.

```{r}

# Call:
# lm(formula = avg_wage ~ treat * time, data = wages2)
# 
# Coefficients:
# (Intercept)        treat         time   treat:time  
#      9.7281       3.5831       0.2719       2.4169

```

What do all the other coefficients represent?

Intercept indicates the expected value for pre-treatment average student wages for the control group (Washington 2016).
treat is the difference between the control group and the treatment group at the pre-treatment time (California 2016 - Washington 2016).
time is the difference between the pre-treatment and post-treatment times for the control group (Washington 2017 - Washington 2016).
Combinations of these coefficients give us back all four means from our dataset.


```{r, message=FALSE}
library(dplyr)
library(ggplot2)
```
# Task 1
```{r}
# read in data
eras <- read.csv("eras_df.csv")
```
# Task 2
```{r}
# print head of dataset
head(eras)

```
# Task 3
```{r}
# line plot for hospital B
hos_b <- eras %>% 
  filter(hospital == "Hospital B") %>%
  ggplot(aes(x = year, y = los)) +
  geom_line() 

hos_b + 
  geom_vline(xintercept = 2015, linetype = "dashed") +
  scale_x_continuous(breaks = c(2010:2019))


  ```
# Task 4
```{r}
# line plot for both hospitals
ggplot(data = eras, aes(x = year, y = los, color = hospital, linetype = hospital))+
  geom_line() +
  geom_vline(xintercept = 2018, linetype = "dashed") +
  scale_x_continuous(breaks = c(2010:2019))


```
# Task 5
```{r}
# restrict data to 2014-2015
eras2 <- eras %>% filter(year %in% c(2014, 2015))

# view dataset
head(eras2)

```
# Task 6
```{r}
# calculate differences within hospitals
Adiff <- 3.705750 - 4.464623
Bdiff <- 2.472485 - 4.026384

print(Adiff)
print(Bdiff)

```
# Task 7
```{r}
# calculate difference in differences
Bdiff - Adiff

```
# Task 8
```{r}
# create indicators
eras2 <- eras2 %>% 
  mutate(treat = ifelse(hospital == "Hospital B", 1, 0)) %>%
  mutate(time = ifelse(year == 2015, 1, 0))

head(eras2)
```
# Task 9
```{r}
# fit linear model with interaction between treat and time
hospital_model <- lm(los ~ treat * time, data = eras2)

# view model
hospital_model

# view model
print(hospital_model)

```
# Task 10
```{r}
# print ATT
hospital_model$coefficients
```
